A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-08-15 simplified_sdc_intro/paper.tex
add some NSE stuff

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling) as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforcing
timestep limiters based on the energy generation or abundance changes
over a timestep.

write these out

\castro\ has the ability to reject a timestep if it detects a failure
and retry with smaller timesteps (subcycling to make up the original
required timestep).  Among the conditions that can trigger this are
density failing below zero during advection, the ODE integration
failing to converge in the implicit solve, or violation of one of the
timestep limiters during the step.  At the moment, this retry
mechanism only works with the CTU hydrodynamics + Strang splitting.
We enable it here, since, as will be shown shortly, this combination
of methods has the most difficulty in integrating the detonation.

@@

The basic idea of spectral deferred corrections is break the timestep
into a number of temporal nodes and use low order difference methods
in a correction equation that, upon iteration, achieves a high-order
solution coupling the different physics together.

\subsection{Simplified-SDC Method}

A second-order accurate time integration method using a simple
deferred correction strategy was shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  A compressible version, for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below:

@@

  \begin{itemize}
  \item Set the first iterate of the primitive-variable reactive source terms from the
    last step:
    \begin{equation}
      \Ic^{n+1/2,(0)}_{\qb} = \Ic^{n-1/2,(\smax)}_{\qb}
    \end{equation}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

  Iterate from $k = 1, \ldots, \smax$.  For second-order accuracy,
  $\smax = 2$ is sufficient.

  \begin{itemize}
  \item {\em Create the Advective Update Term}

@@

    \item predict $\qb$ to the interfaces at $n+1/2$ using the CTU PPM
      method.  The source terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sq = \Sqhydro^{n+1/2} + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      Here we use the iteratively lagged integrals of the primitive variable
      terms accounting only for reactions, $\Ic_\qb^{n+1/2,(k-1)}$, as the
      reactive source.  Any hydrodynamic source terms are time-centered
      using the previous iteration:
      \begin{equation}
        \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
      \end{equation}

    \item solve the Riemann problem at each interface to get a unique
      conserved state on each interface, $\Uc^{n+1/2}_{i+1/2}$

    \item construct the advective update terms, first ignoring the hydrodynamics sources:
      $\Advt{\Uc}^{n+1/2}_{i,j,k}$, e.g., as:
      \begin{align}
        \Advt{\Uc}^{n+1/2,(k)}_{i,j,k} =
          &- \frac{\Fb^\xv(\Uc^{n+1/2}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2}_{i-1/2})}{\Delta x} 
      \end{align}
    Now the conservative hydrodynamics source terms are computed by first updating to the
    new state with advection only\footnote{For a source like gravity, this update can be done first for $\rho$ and then define the new momentum source using $\rho^{\star\star}$ and likewise for energy}, as:
    \begin{equation}
      \Uc^{\star\star} = \Uc^n + \Delta t \Advt{\Uc}^{n+1/2,(k)}
    \end{equation}
    %% first updating the density to the new state with advection only as:
    %% \begin{equation}
    %%   \rho^{\star\star} = \rho^n + \Delta t \Advt{\rho}^{n+1/2,(k)}
    %% \end{equation}
    %% then constructing the momentum source term:
    %% \begin{equation}
    %%   {\bf S}_{\rho\Ub}^{n+1/2} = \frac{1}{2} (\rho^n \gb^n + \rho^{\star\star} \gb^{n+1,(k-1)})
    %% \end{equation}
    %% Then updating the momentum with only advection as:
    %% \begin{equation}
    %%   (\rho \Ub)^{\star\star} = (\rho \Ub)^n + \Delta t \Advt{\rho\Ub}^{n+1/2,(k)} + \Delta t {\bf S}_{\rho\Ub}^{n+1/2}
    %% \end{equation}
    %% and finally constructing the energy source:
    %% \begin{equation}
    %%   S_{\rho E}^{n+1/2} = \frac{1}{2} \left [ (\rho\Ub)^n \cdot \gb^n + (\rho\Ub)^{\star\star} \cdot \gb^{n+1,(k-1)}
    %%     \right ]
    %% \end{equation}

@@

    \begin{itemize}
    \item Define the hydro source terms---these will be time-centered,
      but piecewise-constant in time throughout the integration.

    \item With the spatial discretization done, our system appears as a
      system of ODEs: \MarginPar{what is the right form of acceleration
        here? should we do piecewise linear in time?}
      \begin{equation}
        \left . \odt{\Uc} \right |_{i,j,k} = \Adv{\Uc}^{n+1/2,(k)}_{i,j,k} + \Rb(\Uc)
      \end{equation}

@@

      Note that our advection terms are piecewise constant \MarginPar{need to think about this more}
      approximations, so when we integrate our system, their
      contributions will be linear in time.  Since $(\rho X_k)$ sees a
      linear advective term, $\rho$ does as well.  $(\rho \Ub)$ and
      $(\rho E)$ will also have linear contributions in time from
      advection.  We can see that the internal energy will also have a
      linear contribution by recognizing that the kinetic energy term,
      $K = (\rho |\Ub|^2)/\rho$ is linear in time.
      Since the reactions don't affect $\rho$ and $(\rho \Ub)$, we can
      algebraically update these using this piecewise linear behavior:
      \begin{align}
        \rho(t) &= \rho^n + \Adv{\rho}^{n+1/2,(k)} \, t \\
        (\rho \Ub)(t) &= (\rho \Ub)^n + \Adv{\rho \Ub}^{n+1/2,(k)} \, t
      \end{align}
      At the end of the integration, our new state is $\Uc^{n+1,(k)}$.

@@

\subsection{Nuclear Statistical Equilibrium}

@@

@@

>>>>

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling) as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep:
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le \mathtt{dtnuc\_e}\, \min_{i,j,k} \left\{\frac{e_{i,j,k}}{\dot{e}_{i,j,k}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le \mathtt{dtnuc\_X}\, \min_{i,j,k} \left\{\min_n\frac{X^n_{i,j,k}}{\dot{X}^n_{i,j,k}}\right\}
\end{align}
where $\mathtt{dtnuc\_e}$ and $\mathtt{dtnuc\_X}$ are runtime parameters
used to control the allowed change.

\castro\ has the ability to reject a timestep if it detects a failure
and retry with smaller timesteps (subcycling to make up the original
required timestep).  Among the conditions that can trigger this are
density failing below zero during advection, the ODE integration
failing to converge in the implicit solve, or violation of one of the
timestep limiters during the step.  This means that equations
\ref{eq:dt:nuce} and \ref{eq:dt:nucX} are not reactive, but instead
guaranteed to be met throughout the simulation, because the step is
rejected if they are violated.  The retry mechanism in \castro\ works
with both the Strang and simplified-SDC integration scheme.

@@

The basic idea of spectral deferred corrections is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt
\end{equation}
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  A compressible version, for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

  \begin{itemize}
  \item We need an approximation to how much the reactions alone
    changed the primitive variable state over the timestep, which we
    will denote $\Ic_q$.  Since we do not have any information about
    the current timestep in the first iteration,
    we use the value from the last iteration of the previous timestep:
    \begin{equation}
      \Ic^{n+1/2,(-1)}_{\qb} = \Ic^{n-1/2,(\smax-1)}_{\qb}
    \end{equation}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

  Iterate from $k = 0, \ldots, \smax-1$.  For second-order accuracy,
  $\smax = 2$ is sufficient.  In addition to denoting the time-level
  with a superscript (like $n$ or $n+1$), we'll use a second subscript
  in parentheses to keep track of the iteration.

  \begin{itemize}
  \item {\em Create the advective update term, $\Adv{\Uc}^{n+1/2,(k)}$}

@@

    \item predict $\qb$ to the interfaces at $n+1/2$ using the CTU PPM
      method.  The source terms, $\Sq$, used in the prediction are:
      \begin{equation}
        \Sq = \Sqhydro^{n+1/2} + \Ic_\qb^{n+1/2,(k-1)}
      \end{equation}
      Here we use the iteratively lagged integrals of the primitive variable
      terms accounting only for reactions, $\Ic_\qb^{n+1/2,(k-1)}$, as the
      reactive source.  This is in contrast to Strang-splitting, where no
      explicit reactive source terms are seen by the hydrodynamics update.
      Any hydrodynamic source terms are time-centered
      using the previous iteration:
      \begin{equation}
        \Sqhydro^{n+1/2} = \frac{1}{2} \left ( \Sqhydro^n + \Sqhydro^{n+1,(k-1)} \right )
      \end{equation}

    \item solve the Riemann problem at each interface to get a unique
      conserved state on each interface, $\Uc^{n+1/2,(k)}_{i+1/2}$

    \item construct the advective update terms, first ignoring the hydrodynamics sources:
      $\Advt{\Uc}^{n+1/2,(k)}_{i,j,k}$, e.g., as:
      \begin{align}
        \Advt{\Uc}^{n+1/2,(k)}_{i,j,k} =
          &- \frac{\Fb^\xv(\Uc^{n+1/2,(k)}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2,(k)}_{i-1/2})}{\Delta x}
      \end{align}
    Now the conservative hydrodynamics source terms are computed by first updating to the
    new state with advection only\footnote{For a source like gravity, this update can be done first for $\rho$ and then define the new momentum source using $\rho^{\star\star}$ and likewise for energy}, as:
    \begin{equation}
      \Uc^{\star\star} = \Uc^n + \Delta t \Advt{\Uc}^{n+1/2,(k)}
    \end{equation}
    %% first updating the density to the new state with advection only as:
    %% \begin{equation}
    %%   \rho^{\star\star} = \rho^n + \Delta t \Advt{\rho}^{n+1/2,(k)}
    %% \end{equation}
    %% then constructing the momentum source term:
    %% \begin{equation}
    %%   {\bf S}_{\rho\Ub}^{n+1/2} = \frac{1}{2} (\rho^n \gb^n + \rho^{\star\star} \gb^{n+1,(k-1)})
    %% \end{equation}
    %% Then updating the momentum with only advection as:
    %% \begin{equation}
    %%   (\rho \Ub)^{\star\star} = (\rho \Ub)^n + \Delta t \Advt{\rho\Ub}^{n+1/2,(k)} + \Delta t {\bf S}_{\rho\Ub}^{n+1/2}
    %% \end{equation}
    %% and finally constructing the energy source:
    %% \begin{equation}
    %%   S_{\rho E}^{n+1/2} = \frac{1}{2} \left [ (\rho\Ub)^n \cdot \gb^n + (\rho\Ub)^{\star\star} \cdot \gb^{n+1,(k-1)}
    %%     \right ]
    %% \end{equation}

@@

    We update the state by doing the integral in equation
    \ref{eq:integral:simplesdc}.  Since we are approximating the
    advective term as piecewise constant in time, we can simply use an
    ODE integrator to integrate this, just as we do with the reaction
    system in Strang splitting.  The difference here being that we are
    integrating the conserved variables and the state sees the effect
    of advection as we integrate the reactions.  The ODE form of this
    is
    \begin{equation}
      \odt{\Uc} = \Adv{\Uc}^{n+1,(k)} + \Rb(\Uc)
    \end{equation}

@@

    Our integrator also needs the Jacobian of the system, in terms of
    the conserved variables.  This is different than the form of the
    Jacobian usually used in reaction networks.  We describe the form
    of the Jacobian in appendix \ref{sec:app:jacobian}.

    Note that our advection terms are piecewise constant \MarginPar{need to think about this more}
    approximations, so when we integrate our system, their
    contributions will be linear in time.  Since $(\rho X_k)$ sees a
    linear advective term, $\rho$ does as well.  $(\rho \Ub)$ and
    $(\rho E)$ will also have linear contributions in time from
    advection.  We can see that the internal energy will also have a
    linear contribution by recognizing that the kinetic energy term,
    $K = (\rho |\Ub|^2)/\rho$ is linear in time.
    Since the reactions don't affect $\rho$ and $(\rho \Ub)$, we can
    algebraically update these using this piecewise linear behavior:
    \begin{align}
      \rho(t) &= \rho^n + \Adv{\rho}^{n+1/2,(k)} \, t \\
      (\rho \Ub)(t) &= (\rho \Ub)^n + \Adv{\rho \Ub}^{n+1/2,(k)} \, t
    \end{align}

    At the end of this integration, we have the state at the next iteration,
    $\Uc^{n+1,(k+1)}$

@@

\section{Nuclear Statistical Equilibrium}

For this study we will use the 19 isotope ``aprox19'' network together
with a table that gives the NSE abundances in regions where the system
is in NSE.  The table provides:
\begin{align*}
Y_e &= \sum_k \frac{Z_k X_k}{A_k} \\
\bar{A} &= \left [ \sum_k \frac{X_k}{A_k} \right ]^{-1} \\
\frac{B}{A} &= \sum_k \frac{B_k X_k}{A_k}
\end{align*}
where $B_k$ is the binding energy of nucleus $k$.  In our simulations,
we store these 3 quantities as auxiliary data that is carried along
with the rest of the fluid state in each zone.

For Strang split coupling of hydro and reactions, $DX_k/Dt = 0$,
and our evolution equations are:
\begin{align*}
\frac{DY_e}{Dt} &= \sum_k \frac{Z_k}{A_k} \frac{DX_k}{Dt} = 0 \\
\frac{D}{Dt} \frac{1}{\bar{A}} &= - \frac{1}{\bar{A}^2} \frac{D\bar{A}}{Dt} = \sum_k \frac{1}{A_k} \frac{DX_k}{Dt} = 0 \rightarrow \frac{D\bar{A}}{Dt} = 0 \\
\frac{D}{Dt} \left (\frac{B}{A} \right ) &= \sum_k \frac{B_k}{A_k} \frac{DX_k}{Dt} = 0
\end{align*}
Therefore each of these auxillar equations obeys an advection
equation in the hydro part of the advancement.  In the SDC algorithm,
there will be a reactive source (an $\Ic_q$) for each of these that is
computed in the same manner as above.

The NSE table was generated using a 125 nuclei reaction network and is
the same as described in \cite{ma:2013}.  The compositional quantities
it carries, $\bar{A}$ and $Y_e$ and not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always take the composition quantities in
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly. Our equation of state needs $\bar{Z}$ in addition
to the auxiliary quantities, which is is easily computed as
\begin{equation}
\bar{Z} = \bar{A} Y_e
\end{equation}

For Strang splitting, the basic flow of using the aprox19 + NSE network
is as follows:
\begin{itemize}
\item initialize the problem, including $X_k$

\item fill the initial aux data with $Y_e$, $\bar{A}$, and $(B/A)$.

\item in hydro, update these quantities simply via advection

\item for the reactive update:

  \begin{itemize}
  \item check if NSE applies (see below)

  \item if we are in an NSE region:

    \begin{itemize}
    \item use $\rho$, $T$, and $Y_e$ to call the table. This
      returns: $dY_e/dt$, $(B/A)_{\rm out}$, and $\bar{A}_{\rm out}$.

    \item update $Y_e$:
    \begin{equation}
      (Y_e)_{\rm out} = (Y_e)_{\rm in} - \Delta t \frac{dY_e}{dt}
    \end{equation}

  \item update the auxiliary state value of $\bar{A}$ to be simply
    $\bar{A}_{\rm out}$ from the table.

  \item compute the energy generation rate, $\Sdot$ as:
    \begin{equation}
      \Sdot = \eta \left [ \left ( \frac{B}{A} \right )_{\rm out} -
        \left ( \frac{B}{A} \right )_{\rm in} \right ] N_A \frac{1}{\Delta t}
    \end{equation}
    where $\eta$ is an inertia term < 1 to prevent the energy changing
    too much in one set.

  \item set the auxiliary state value of $(B/A)$ as:
    \begin{equation}
      \left (\frac{B}{A}\right )_{\rm out} = \left (\frac{B}{A}\right )_{\rm in} + \eta \left [ \left ( \frac{B}{A}\right )_{\rm out} - \left (\frac{B}{A} \right )_{\rm in} \right]
    \end{equation}

  \item update the mass fractions, $X_k$, using the values from the table
    \end{itemize}

  \item if we are not in NSE:

    \begin{itemize}
    \item integrate the aprox19 network as usual
    \item update the aux quantities at the end of the burn
    \end{itemize}
  \end{itemize}
\end{itemize}

We use the following conditions to determine if we are in NSE:
\begin{itemize}
\item $\rho > \rho_\mathrm{nse}$
\item $T > T_\mathrm{nse}$
\item $X(\isotm{C}{12}) < A_\mathrm{nse}$
\item $X(\isotm{He}{4}) + X(\isotm{Cr}{48}) + X(\isotm{Fe}{52}) + X(\isotm{Fe}{54}) + X(\isotm{Ni}{56}) > B_\mathrm{nse}$
\end{itemize}

@@

\subsection{Reacting convergence test problem}

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  We run that same test here with the
simplified SDC algorithm.

@@

\label{sec:app:jacobian}
